---
title: "TvS-AlgalBloom-EWS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TvS-AlgalBloom-EWS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Rmarkdown to create all analyses and figures in *Evaluating the performance of temporal and spatial early warning statistics of algal blooms* by CD Buelo, ML Pace, SR Carpenter, EH Stanley, DA Ortiz, DT Ha.

```{r libraries, warning=FALSE, message=FALSE}
library(devtools)
# devtools::install_github("cbuelo/tvsews") 
# or install.packages('your/path/to/tvsews_0.1.0.tar.gz', repos=NULL)
library(tvsews)
library(tidyverse)
library(sf)
library(viridis)
library(gridExtra)
library(cowplot)
library(grid)
library(broom)
library(nlme)
```
# Plot state variable time series (Figure 1)
```{r Figure1, warning=FALSE, echo=TRUE, message=FALSE, fig.width=9, fig.height=9}
plot_state_vars(
  time_series = ts_data %>% filter(Year >= 2018),
  bloom_fert_df = bloom_fert_dates %>% filter(Year >= 2018 & Lake == "R")
) +
  guides(linetype= FALSE) +
  theme(
    legend.position=c(0.1, 0.93), 
    legend.background = element_blank(), 
    legend.key = element_blank(),
    axis.text=element_text(size=12),
    axis.title=element_text(size=16),
    strip.text = element_text(size=16)
    )
```

# Plot example spatial maps (Figure 2)

```{r Figure2, warning=FALSE, message=FALSE, echo=TRUE, display=FALSE, fig.width=9, fig.height=9}
plot_spatial_data(
  spatial_data = flame_data,
  samples_to_plot = data.frame(Year = 2019, DOY = c(165, 210, 228), stringsAsFactors = FALSE)
)
```

## Run rolling window analyses for current experiment
```{r, echo=TRUE, warning=FALSE, message=FALSE}
# define variables to use
use_vars = c("Manual_Chl", "BGA_HYLB", "DO_Sat", "pH")
# calculate rolling window statistics
rw_stats = calc_rolling_stats(ts_data %>% filter(Year >= 2018),
                              var_cols=use_vars)
```

# Run QD method for current experiment
```{r, warnings=FALSE, message=FALSE}
qd_s = run_qd(rw_stats, var_cols=use_vars, ar1_alarm_rho = 0.95)
qd_a = format_qd(qd_s, bloom_fert_df = bloom_fert_dates)
```

# Plot the rolling window time series and QD alarms (Figure 3)

```{r Figure3, warning=FALSE, echo=TRUE, message=FALSE, fig.width=12, fig.height=8}
f3 = plot_temporal_EWS(rolling_window_stats = rw_stats, 
               qd_alarms = qd_a, 
               bloom_fert_df = bloom_fert_dates %>% filter(Year == 2019 & Lake == "R")
               ) 
```

# Calculate spatial statistics

```{r, warning=FALSE, message=FALSE}
spatial_stats <- calc_spatial_stats(
  spatial_data = flame_data,
  var_cols = c("BGApc_ugL_tau", "ODO_percent_tau", "pH_tau")
)
```

# Plot time series of spatial statistics (Figure 4)
```{r Figure4, echo=TRUE, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
plot_spatial_EWS(spatial_stats, 
          var_rename_vec = c(`Phyco` = "BGApc_ugL_tau", 
                             `D.O. sat.` = "ODO_percent_tau", 
                             pH = "pH_tau"))
```
# Calculate rolling window stats and QD alarms for all data

```{r, echo=TRUE, warning=FALSE, message=FALSE}
rw_stats_all = calc_rolling_stats(ts_data, var_cols=use_vars)
qd_s_all = invisible(run_qd(rw_stats_all, var_cols=use_vars, ar1_alarm_rho = 0.95))
qd_a_all = format_qd(qd_s_all, bloom_fert_df = bloom_fert_dates)
qd_rates_all = calc_alarm_rates(qd_a_all)
```

# Plot the true and false positive alarm rates (Figure 5)

```{r Figure5, fig.width=10, fig.height=6, echo=TRUE}
plot_alarm_rates(qd_rates_all, plot_title="")
```

# Calculate info for table on statistic trends 

## Temporal slope trends
```{r, echo=TRUE, warning=FALSE, message=FALSE}
rw_slope_0 = rw_stats %>% 
  filter(Stat %in% c("SD", "Ar1")) %>% 
  select(Year, DOYtrunc, Stat, Lake, Width, all_of(use_vars)) %>% 
  pivot_longer(cols=use_vars, names_to="Variable") %>% 
  mutate(LakeType = ifelse(Lake == "L", "Reference", "Experimental"))

rw_slope = rw_slope_0 %>% 
  select(-Lake) %>% 
  pivot_wider(names_from=LakeType, values_from = value) %>% 
  filter(!is.na(Reference) & !is.na(Experimental))

# combine, filter, and calculate difference
rw_slope_data = rw_slope %>% 
  mutate(Stat_Diff = Experimental - Reference)

# add fert and bloom start dates
rw_slope_data_dates = rw_slope_data %>%  
  left_join(bloom_fert_dates %>% filter(Year >= 2018 & Lake == "R"), by="Year") %>% 
  select(-Lake) %>% 
  mutate(fertStartDOY = ifelse(is.na(fertStartDOY), 367, fertStartDOY)) %>% 
  mutate(fertEndDOY = ifelse(is.na(fertEndDOY), 368, fertEndDOY)) %>% 
  mutate(bloomStartDOY = ifelse(is.na(bloomStartDOY), 367, bloomStartDOY))

# classify different time periods based on fert/bloom start dates
rw_slope_data_dates$TimePeriod = "NA"
rw_slope_data_dates = rw_slope_data_dates %>% 
  mutate(TimePeriod = ifelse(
    rw_slope_data_dates$DOYtrunc < rw_slope_data_dates$fertStartDOY, 
    "preFert", 
    TimePeriod))
rw_slope_data_dates = rw_slope_data_dates %>% 
  mutate(TimePeriod = ifelse(rw_slope_data_dates$DOYtrunc >= rw_slope_data_dates$fertStartDOY &
                               rw_slope_data_dates$DOYtrunc < rw_slope_data_dates$bloomStartDOY, 
         "alarmPeriod",
         TimePeriod))
rw_slope_data_dates = rw_slope_data_dates %>% 
  mutate(TimePeriod = ifelse(rw_slope_data_dates$DOYtrunc >= rw_slope_data_dates$bloomStartDOY, 
         "postBloom",
         TimePeriod))

# filter just alarm period after fertilization began but before bloom
rw_slope_alarmPeriod = rw_slope_data_dates %>% 
  filter(TimePeriod == "alarmPeriod")

# do Kendall's tau test, just get variables that have p value < 0.05
rw_slopes_significant = rw_slope_alarmPeriod %>% 
  group_by(Stat, Variable) %>% 
  do(DOYfit = tidy(cor.test(.$DOYtrunc, .$Stat_Diff, data = ., method="kendall"))) %>%
  unnest(DOYfit) %>%
  # filter(term == "DOYtrunc") %>%
  filter(Stat %in% c("SD", "Ar1")) %>% 
  filter(p.value < 0.05)

rw_slopes_significant
```

## Spatial slope trends
```{r, warning=FALSE, echo=TRUE, message=FALSE}
# add the fert and bloom dates to the spatial stats
spatial_stats_slope_0 = spatial_stats %>% 
  mutate(LakeType = ifelse(Lake == "L", "Reference", "Experimental"))

spatial_stats_slope = spatial_stats_slope_0 %>% 
  select(-Lake) %>% 
  pivot_wider(names_from=LakeType, values_from = Value) %>% 
  filter(!is.na(Reference) & !is.na(Experimental))

# combine, filter, and calculate difference
spatial_stats_slope_data = spatial_stats_slope %>% 
  mutate(Stat_Diff = Experimental - Reference)

# add fert and bloom start dates
spatial_stats_slope_data_dates = spatial_stats_slope_data %>%  
  left_join(bloom_fert_dates %>% filter(Year >= 2018 & Lake == "R"), by="Year") %>% 
  select(-Lake) %>% 
  mutate(fertStartDOY = ifelse(is.na(fertStartDOY), 367, fertStartDOY)) %>% 
  mutate(fertEndDOY = ifelse(is.na(fertEndDOY), 368, fertEndDOY)) %>% 
  mutate(bloomStartDOY = ifelse(is.na(bloomStartDOY), 367, bloomStartDOY))

# classify different time periods based on fert/bloom start dates
spatial_stats_slope_data_dates$TimePeriod = "NA"
spatial_stats_slope_data_dates = spatial_stats_slope_data_dates %>% 
  mutate(TimePeriod = ifelse(
    spatial_stats_slope_data_dates$DOY < spatial_stats_slope_data_dates$fertStartDOY, 
    "preFert",
    TimePeriod))
spatial_stats_slope_data_dates = spatial_stats_slope_data_dates %>% 
  mutate(TimePeriod = ifelse(spatial_stats_slope_data_dates$DOY >= spatial_stats_slope_data_dates$fertStartDOY &
                               spatial_stats_slope_data_dates$DOY < spatial_stats_slope_data_dates$bloomStartDOY, 
         "alarmPeriod",
         TimePeriod))
spatial_stats_slope_data_dates = spatial_stats_slope_data_dates %>% 
  mutate(TimePeriod = ifelse(spatial_stats_slope_data_dates$DOY >= spatial_stats_slope_data_dates$bloomStartDOY, 
         "postBloom",
         TimePeriod))

spatial_slope_results = spatial_stats_slope_data_dates %>% 
  filter(TimePeriod == "alarmPeriod") %>%
  group_by(Stat, Variable) %>% 
  do(DOYfit = tidy(cor.test(.$DOY, .$Stat_Diff, data = ., method="kendall"))) %>% 
  unnest(DOYfit) %>% 
  arrange(Stat, Variable) %>% 
  filter(p.value < 0.05)

spatial_slope_results
```
## Paired t-test if mean difference between lakes during alarm period is different from 0
```{r}
spatial_means_results = spatial_stats_slope_data_dates %>%
  filter(TimePeriod == "alarmPeriod") %>%
  group_by(Stat, Variable) %>% 
  do(tidy(t.test(x=.$Experimental, y=.$Reference, paired=TRUE))) %>% 
  filter(p.value < 0.05)

spatial_means_results
```

# Mixed model to account for non-independence
```{r}
# look at difference between lakes for spatial EWS
hold = spatial_stats_slope_data_dates %>%
  # run for all of BGApc_ugL_tau, ODO_percent_tau, and pH_tau, and Stat = SD, Moran's I
  filter(TimePeriod == "alarmPeriod" & Variable == "pH_tau" & Stat == "Moran's I") %>% 
  select(DOY, Stat_Diff, Stat, Variable) %>% 
  arrange(DOY) %>% 
  mutate(sample = row_number())

hold_gls = gls(Stat_Diff ~ 1, data=hold, correlation = corAR1(form=~sample))
summary(hold_gls)
```



# Skewness and kurtosis
```{r, warning=FALSE}
# calculate RW and spatial skewness and kurtosis
use_vars = c("Manual_Chl", "BGA_HYLB", "DO_Sat", "pH")
ts_sk = calc_rolling_stats(ts_data %>% filter(Year >= 2018),
                           var_cols=use_vars,
                           statistics = c("Skewness", "Kurtosis"))

bloom_dates <- bloom_fert_dates %>%
  tidyr::pivot_longer(cols = c("fertStartDOY", "fertEndDOY", "bloomStartDOY"), 
                      names_to = "Start Of", 
                      values_to = "DOYtrunc") %>%
  mutate(
    `Start Of` = ifelse(`Start Of` == "fertStartDOY", "nutrients", `Start Of`),
    `Start Of` = ifelse(`Start Of` == "bloomStartDOY", "bloom", `Start Of`)
  ) %>%
  filter(`Start Of` %in% c("nutrients", "bloom") & Lake == "R" & Year >= 2018)

sp_sk = calc_spatial_stats(statistics = c("skew", "kurt"))
```
## plot temporal skew and kurt
```{r, fig.width=12, fig.height=6}
# plot the time series
ts_long = ts_sk %>%
  pivot_longer(cols = c("Manual_Chl", "BGA_HYLB", "DO_Sat", "pH")) %>%
  mutate(name = factor(name, levels = c("Manual_Chl", "BGA_HYLB", "DO_Sat", "pH"),
                       labels = c("Chl-a", "Phyco", "D.O. Sat", "pH"), ordered = T))

ts_skewplot = ggplot(ts_long %>% filter(Stat == "Skewness"),
                     aes(x=DOYtrunc, y=value, color=Lake)) +
  geom_line(size=1.5) +
  facet_grid(rows=vars(name), cols=vars(Year), scales="free_y") +
  theme_bw() +
  scale_color_manual(values = c("R" = "firebrick3", "L" = "royalblue3"), guide="none") +
  geom_vline(data = bloom_dates, aes(xintercept = DOYtrunc, linetype = `Start Of`)) +
  scale_linetype_manual(breaks = c("nutrients", "bloom"), values = c("dashed", "solid"), guide = FALSE) +
  ggtitle("Skewness") +
  theme(plot.title = element_text(hjust=0.5)) +
  labs(x="Day Of Year")

ts_kurtplot = ggplot(ts_long %>% filter(Stat == "Kurtosis"),
                     aes(x=DOYtrunc, y=value, color=Lake)) +
  geom_line(size=1.5) +
  facet_grid(rows=vars(name), cols=vars(Year), scales="free_y") +
  theme_bw() +
  scale_color_manual(values = c("R" = "firebrick3", "L" = "royalblue3"), guide = "none") +
  geom_vline(data = bloom_dates, aes(xintercept = DOYtrunc, linetype = `Start Of`)) +
  scale_linetype_manual(breaks = c("nutrients", "bloom"), values = c("dashed", "solid"), guide = FALSE) +
  ggtitle("Kurtosis") +
  theme(plot.title = element_text(hjust=0.5)) +
  labs(x="Day Of Year")

ts_plot = gridExtra::grid.arrange(ts_skewplot, ts_kurtplot, nrow=1)
# ggsave("~/Documents/Cascade/SpatialSquealAnalysis/Figures/RW_skewness_kurtosis_v1.png", ts_plot, )
```

## plot spatial skew and kurt
```{r, fig.width=12, fig.height=6}
sp_skewplot = sp_sk %>%
  mutate(Variable = factor(Variable, 
                           levels = c("BGApc_ugL_tau", "ODO_percent_tau", "pH_tau"), 
                           labels = c("Phyco", "D.O. Sat", "pH"), ordered = T)) %>%
  filter(Stat == "skew") %>%
  ggplot(aes(x=DOY, y=Value, color=Lake)) +
  geom_line(size=1.5) +
  facet_grid(rows=vars(Variable), cols=vars(Year), scales="free_y") +
  theme_bw() +
  scale_color_manual(values = c("R" = "firebrick3", "L" = "royalblue3"), guide="none") +
  geom_vline(data = bloom_dates, aes(xintercept = DOYtrunc, linetype = `Start Of`)) +
  scale_linetype_manual(breaks = c("nutrients", "bloom"), values = c("dashed", "solid"), guide = FALSE) +
  ggtitle("Skewness") +
  theme(plot.title = element_text(hjust=0.5)) +
  labs(x="Day Of Year")

sp_kurtplot = sp_sk %>%
  mutate(Variable = factor(Variable, 
                           levels = c("BGApc_ugL_tau", "ODO_percent_tau", "pH_tau"), 
                           labels = c("Phyco", "D.O. Sat", "pH"), ordered = T)) %>%
  filter(Stat == "kurt") %>%
  ggplot(aes(x=DOY, y=Value, color=Lake)) +
  geom_line(size=1.5) +
  facet_grid(rows=vars(Variable), cols=vars(Year), scales="free_y") +
  theme_bw() +
  scale_color_manual(values = c("R" = "firebrick3", "L" = "royalblue3"), guide="none") +
  geom_vline(data = bloom_dates, aes(xintercept = DOYtrunc, linetype = `Start Of`)) +
  scale_linetype_manual(breaks = c("nutrients", "bloom"), values = c("dashed", "solid"), guide = FALSE) +
  ggtitle("Kurtosis") +
  theme(plot.title = element_text(hjust=0.5)) +
  labs(x="Day Of Year")

sp_plot = gridExtra::grid.arrange(sp_skewplot, sp_kurtplot, nrow=1)
```

## time series slopes and mean diffs setup
```{r}
# calculate slopes and differences between lakes
rw_slope_0 = ts_sk %>% 
  select(Year, DOYtrunc, Stat, Lake, Width, all_of(use_vars)) %>% 
  pivot_longer(cols=use_vars, names_to="Variable") %>% 
  mutate(LakeType = ifelse(Lake == "L", "Reference", "Experimental"))

rw_slope = rw_slope_0 %>% 
  select(-Lake) %>% 
  pivot_wider(names_from=LakeType, values_from = value) %>% 
  filter(!is.na(Reference) & !is.na(Experimental))

# combine, filter, and calculate difference
rw_slope_data = rw_slope %>% 
  mutate(Stat_Diff = Experimental - Reference)

# add fert and bloom start dates
rw_slope_data_dates = rw_slope_data %>%  
  left_join(bloom_fert_dates %>% filter(Year >= 2018 & Lake == "R"), by="Year") %>% 
  select(-Lake) %>% 
  mutate(fertStartDOY = ifelse(is.na(fertStartDOY), 367, fertStartDOY)) %>% 
  mutate(fertEndDOY = ifelse(is.na(fertEndDOY), 368, fertEndDOY)) %>% 
  mutate(bloomStartDOY = ifelse(is.na(bloomStartDOY), 367, bloomStartDOY))

# classify different time periods based on fert/bloom start dates
rw_slope_data_dates$TimePeriod = "NA"
rw_slope_data_dates = rw_slope_data_dates %>% 
  mutate(TimePeriod = ifelse(
    rw_slope_data_dates$DOYtrunc < rw_slope_data_dates$fertStartDOY, 
    "preFert", 
    TimePeriod))
rw_slope_data_dates = rw_slope_data_dates %>% 
  mutate(TimePeriod = ifelse(rw_slope_data_dates$DOYtrunc >= rw_slope_data_dates$fertStartDOY &
                               rw_slope_data_dates$DOYtrunc < rw_slope_data_dates$bloomStartDOY, 
                             "alarmPeriod",
                             TimePeriod))
rw_slope_data_dates = rw_slope_data_dates %>% 
  mutate(TimePeriod = ifelse(rw_slope_data_dates$DOYtrunc >= rw_slope_data_dates$bloomStartDOY, 
                             "postBloom",
                             TimePeriod))
```

## significant slopes
```{r}
# filter just alarm period after fertilization began but before bloom
rw_slope_alarmPeriod = rw_slope_data_dates %>% 
  filter(TimePeriod == "alarmPeriod")

# do Kendall's tau test, just get variables that have p value < 0.05
rw_slopes_significant = rw_slope_alarmPeriod %>% 
  group_by(Stat, Variable) %>% 
  do(DOYfit = tidy(cor.test(.$DOYtrunc, .$Stat_Diff, data = ., method="kendall"))) %>%
  unnest(DOYfit) %>%
  filter(p.value < 0.05)

rw_slopes_significant
```

## significant spatial skew and kurtosis slopes
```{r}
spatial_stats_slope_0 = sp_sk %>% 
  mutate(LakeType = ifelse(Lake == "L", "Reference", "Experimental"))

spatial_stats_slope = spatial_stats_slope_0 %>% 
  select(-Lake) %>% 
  pivot_wider(names_from=LakeType, values_from = Value) %>% 
  filter(!is.na(Reference) & !is.na(Experimental))

# combine, filter, and calculate difference
spatial_stats_slope_data = spatial_stats_slope %>% 
  mutate(Stat_Diff = Experimental - Reference)

# add fert and bloom start dates
spatial_stats_slope_data_dates = spatial_stats_slope_data %>%  
  left_join(bloom_fert_dates %>% filter(Year >= 2018 & Lake == "R"), by="Year") %>% 
  select(-Lake) %>% 
  mutate(fertStartDOY = ifelse(is.na(fertStartDOY), 367, fertStartDOY)) %>% 
  mutate(fertEndDOY = ifelse(is.na(fertEndDOY), 368, fertEndDOY)) %>% 
  mutate(bloomStartDOY = ifelse(is.na(bloomStartDOY), 367, bloomStartDOY))

# classify different time periods based on fert/bloom start dates
spatial_stats_slope_data_dates$TimePeriod = "NA"
spatial_stats_slope_data_dates = spatial_stats_slope_data_dates %>% 
  mutate(TimePeriod = ifelse(
    spatial_stats_slope_data_dates$DOY < spatial_stats_slope_data_dates$fertStartDOY, 
    "preFert", 
    TimePeriod))
spatial_stats_slope_data_dates = spatial_stats_slope_data_dates %>% 
  mutate(TimePeriod = ifelse(
    spatial_stats_slope_data_dates$DOY >= spatial_stats_slope_data_dates$fertStartDOY &
      spatial_stats_slope_data_dates$DOY < spatial_stats_slope_data_dates$bloomStartDOY, 
                             "alarmPeriod",
                             TimePeriod))
spatial_stats_slope_data_dates = spatial_stats_slope_data_dates %>% 
  mutate(TimePeriod = ifelse(
    spatial_stats_slope_data_dates$DOY >= spatial_stats_slope_data_dates$bloomStartDOY, 
                             "postBloom",
                             TimePeriod))

spatial_slope_results = spatial_stats_slope_data_dates %>% 
  filter(TimePeriod == "alarmPeriod") %>%
  group_by(Stat, Variable) %>% 
  do(DOYfit = tidy(cor.test(.$DOY, .$Stat_Diff, data = ., method="kendall"))) %>% 
  unnest(DOYfit) %>% 
  arrange(Stat, Variable) %>% 
  filter(p.value < 0.05)

spatial_slope_results
```


```{r}
# look at difference between lakes for spatial EWS
hold = spatial_stats_slope_data_dates %>% 
  # run for all of BGApc_ugL_tau, ODO_percent_tau, and pH_tau, and Stat = kurt, skew
  filter(TimePeriod == "alarmPeriod" & Variable == "pH_tau" & Stat == "kurt") %>%
  select(DOY, Stat_Diff) %>% 
  arrange(DOY) %>% 
  mutate(sample = row_number())

hold_gls = gls(Stat_Diff ~ 1, data=hold, correlation = corAR1(form=~sample))
summary(hold_gls)

# Significant: 

## ODO skew: p = 0, difference is NEGATIVE
## ODO kurt: p = 2e-4, difference is POSITIVE

## pH skewness: p = 0, difference is NEGATIVE
## pH kurtosis: p = 0.021, difference is POSITIVE

```
